Appendix 1: sdmTMB statistical model description

### 2022-01-06 doc/appendix-model.Rmd
Lewis edits

* $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}_{1}$, $\bm{X}_{2}$, and $\bm{X}_{3}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $O_{\s,t}$ represents an offset: a covariate (usually log transformed) with a coefficient fixed at one;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; 
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\s,t}$ represents a spatiotemporal component (a random field), $\epsilon_{\s,t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

@@

$$
s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln\ \left(19\right) \cdot \left( x-s50 \right)
     / \left(s95 - s50 \right) } \right] }^{-1},
$$

where $s$ represents the logistic function, 
$\psi$ is a scaling parameter (controlling the height of the y-axis for the response; unconstrained), 
$\tau$ is an intercept, 
$s50$ is a parameter controlling the point at which the function reaches 50% of the maximum ($\psi$), 
and $s95$ is a parameter controlling the point at which the function reaches 95%. 
The parameter $s50$ is unconstrained but $s95$ is constrained to be larger than $s50$.

@@

Similarly to the spatial random fields, these spatiotemporal random fields (and all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$):

$$
\boldsymbol{\epsilon_{t}} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right)
$$

@@

@@

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \gamma_{t} x_{\s,t} + \ldots \right),\\
  \gamma_{t=1} &\sim \operatorname{Uniform} \left(-\infty, \infty \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\gamma_{t-1}, \sigma^2_{\gamma} \right),
\end{aligned}
$$

@@

where $\alpha_g$ is an example optional "random" intercept---an intercept
with mean zero that varies by level $g$ and is constrained by $\sigma_\alpha$. 
This is defined by the `formula` argument via the `(1 | g)` syntax as in lme4 or glmmTMB. 
There can be multiple random intercepts, despite only showing one above.
E.g., `(1 | g) + (1 | h)`, in which case they are assumed independent and uncorrelated from each other.

@@

$$
\operatorname{Binomial} \left(N, \mu \right)
$$
where $N$, is the size or number of trials, and $\mu$ is the probability of success for each trial.
If $N = 1$, the distribution becomes the Bernoulli distribution.
Internally, the distribution is parameterized as the [robust
version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
<!-- EW: Put lme4 and glmmTMB in '' too?-->
Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:

@@

$$
\operatorname{Beta} \left(\mu \phi, (1 - \mu) \phi \right)
$$
<!-- EW: betareg in ''?-->
where $\mu$ is the mean and $\phi$ is a precision parameter.
This parametrization follows @ferrari2004 and the betareg R package [@cribari-neto2010].
The variance is $\mu (1 - \mu) / (\phi + 1)$.

@@

where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 2, and $\phi$ is the dispersion parameter.
The Tweedie distribution can be helpful for modelling data that positive and continuous but also contains zeros.

@@

## Matern parameterization

@@

TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longitude ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)). Anisotropy can be turned on and off with the logical `anisotropy` argument to `sdmTMB()`. There are a number of ways to implement anisotropic covariance [@fuglstad2015a], and we adopt a 2-parameter rotation matrix $\textbf{H}$. The elements of $\textbf{H}$ are defined by the parameter vector $\textbf{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$.

@@

In some cases the spatial domain of interest may be complex and bordered by land or water (e.g., coastlines, islands, lakes). 
SPDE models allow for barriers to be included via barrier meshes [@bakka2019]. 
With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`. 
The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers. 
See `?sdmTMB::add_barrier_mesh` for examples.

>>>>

* $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}_{1}$, $\bm{X}_{2}$, and $\bm{X}_{3}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $O_{\s,t}$ represents an offset: a covariate (usually log transformed) with a coefficient fixed at one;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; 
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a spatial component (a random field), $\omega_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\s,t}$ represents a spatiotemporal component (a random field), $\epsilon_{\s,t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

@@

$$
s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln \left(19\right) \cdot \left( x-s50 \right)
     / \left(s95 - s50 \right) } \right] }^{-1},
$$

where $s$ represents the logistic function, 
$\psi$ is a scaling parameter (controlling the height of the y-axis for the response; unconstrained), 
$\tau$ is an intercept, 
$s50$ is a parameter controlling the point at which the function reaches 50% of the maximum ($\psi$), 
and $s95$ is a parameter controlling the point at which the function reaches 95% of the maximum. 
The parameter $s50$ is unconstrained but $s95$ is constrained to be larger than $s50$.

@@

Similarly to the spatial random fields, these spatiotemporal random fields (including all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$):

$$
\boldsymbol{\epsilon_{t}} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right).
$$

@@

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \gamma_{t} \bm{X}_{2,s,t} + \ldots \right),\\
  \gamma_{t=1} &\sim \operatorname{Uniform} \left(-\infty, \infty \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\gamma_{t-1}, \sigma^2_{\gamma} \right),
\end{aligned}
$$

@@

@@

where $\alpha_g$ is an example optional "random" intercept---an intercept
with mean zero that varies by level $g$ and is constrained by $\sigma_\alpha$. 
This is defined by the `formula` argument via the `(1 | g)` syntax as in lme4 or glmmTMB. 
There can be multiple random intercepts, despite only showing one above.
E.g., `(1 | g1) + (1 | g2)`, in which case they are assumed independent and uncorrelated from each other.

@@

$$
\operatorname{Binomial} \left(N, \mu \right)
$$
where $N$ is the size or number of trials, and $\mu$ is the probability of success for each trial.
If $N = 1$, the distribution becomes the Bernoulli distribution.
Internally, the distribution is parameterized as the [robust
version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:

@@

$$
\operatorname{Beta} \left(\mu \phi, (1 - \mu) \phi \right)
$$
where $\mu$ is the mean and $\phi$ is a precision parameter.
This parametrization follows @ferrari2004 and the betareg R package [@cribari-neto2010].
The variance is $\mu (1 - \mu) / (\phi + 1)$.

@@

where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 2, and $\phi$ is the dispersion parameter.
The Tweedie distribution can be helpful for modelling data that are positive and continuous but also contains zeros.

@@

## Matérn parameterization

@@

TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longitude ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)). Anisotropy can be turned on and off with the logical `anisotropy` argument to `sdmTMB()`. There are a number of ways to implement anisotropic covariance [@fuglstad2015a], and we adopt a 2-parameter rotation matrix $\textbf{H}$. The elements of $\textbf{H}$ are defined by the parameter vector $\boldsymbol{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$.

@@

In some cases the spatial domain of interest may be complex and bounded by some barrier such as by land or water (e.g., coastlines, islands, lakes). 
SPDE models allow for barriers to be included via barrier meshes [@bakka2019]. 
With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`. 
The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers. 
See `?sdmTMB::add_barrier_mesh` for examples.
